Generalized Bradley-Terry models

### 2025-03-06 inst/jss_hyper3_real.Rnw
more reworking

One generalization of the Bradley-Terry model is due
to~\citet{luce1959}, in which the probability of competitor~$i$
winning in a field of~$j=1,\ldots, n$ is $\frac{p_i}{p_1+\cdots
  +p_n}$.  It is then natural to quantify some ephemeral enhancement
to competitor~$i$ using a multiplicative factor $\lambda$.  Then
$\Prob(\mbox{$i$ wins})=\frac{\lambda p_i}{p_1+\cdots+\lambda
  p_i+\cdots+p_n}$.  Examples of such enhancements include the
home-ground advantage in football and rugby, playing white in chess,
or XXXX.  Note that $\lambda < 1$ would correspond to a decrease in
competitive ability.

Further, one might follow HANKIN and consider teams of players.  In
doubles tennis, for example, $\left\lbrace p_1,p_2\right\rbrace$ might
play $\left\lbrace p_3,p_4\right\rbrace$ and win with probability
$\frac{p_1+p_2}{p_1+p_2+p_3+p_4}$.  It is natural to combine this idea
with that of a multiplicative factor.  For example, suppose $p_1$ has
a special move against $p_3$, and $p_4$ has just been given a
performance-enhancing drug, we might have a likelihood functions
proportional to $\frac{\lambda p_1+ p_2}{p_1+p_2+p_3+\mu p_4}$.  Then
the effectiveness of $p_3$'s special move and the enhancement afforded
by the drug are measured by $\lambda$ and $\mu$ respectively with
values $>1$ corresponding to a positive effect.

@@

In {\tt hyper3} idiom, a term such as $\lambda p_R + p_C$ is
represented as a named numeric vector.  For example, if the home
strength were $\lambda = 1.8$ then the first bracket would be {\tt
  c(Rangers=1.8, Celtic=1)} with an associated power of $-81$.  The
computational methodology is a generalization of that used
in~\cite{hankin2022} except that the multiplicative factors are real
(as opposed to non-negative integer), and also unknown.

@@

Further, we may calculate a support function over the
$(\lambda,D)$-plane, shown in figure~\ref{contourlambdaD}.  For
example, we may test the hypothesis that $\lambda=1$ [corresponding to
  the right hand side of figure~\ref{contourlambdaD}] for which the
support is about \Sexpr{round(f(maxv) - jj$objective,2)}, far
exceeding Edwards's two units of support per degree of freedom
criteron.  Alternatively, we can observe that this lies in the tail
region of its asymptotic null $\chi^2_4$ distribution with a $p$-value
of about \Sexpr{signif(pchisq(f(maxv) -
  jj$objective,lower.tail=FALSE,df=2),digits=2)}.

@@

The probability model of Figure~\ref{tikzabcde} is applied, in a
simple one-parameter form, to two examples: Formula 1 motor racing,
and also a preference rating.

@@

The red bus-blue bus problem was originally stated by
\cite{chipman1960} and discussed by \cite{davidson1970} and
\cite[p153]{alvo2014}.  was \cite{debreu1960}.  The Holiday problem is
discussed in \cite{tverskey1972}.  A traveller has a choice of going
to work by different transport methods.  Here, we can assume that the
traveller has strong but unknown views on different forms of transport
which might incorporate a random component.  If considering a car or
bus, the colour of the bus is of no importance to him.  However, when
offered the choice between a red bus or a blue bus, the colour becomes
important (having nothing else to guide his decision).  Now, he does
not really care what colour the bus is.  If we ask him to rank his
options, it is highly probable that he will put RB and BB
consecutively (because they are essentially indistinguishable).  Can
we quantify the strength of this effect?  To do this, we define a
bespoke function \code{RB\_BB\_LF()} which returns a \code{hyper3}
log-likelihood function corresponding to repeated observations of our
commuter's reported ranks for the five options:

@@

Easily satisfying Edwards's two-units-of-support criterion;
alternatively, we could observe that this observation lies in the tail
region of its asymptotic $\chi^2_1$ distribution (Wilks's theorem),
with a $p$-value of \Sexpr{signif(pchisq(suppdiff*2, df=1,
  lower.tail=FALSE),3)}.

@@

To assess this, we follow exactly the same protocol as for the sushi
preference dataset but with modified Plackett-Luce likelihoods using
the deathorder statistic from the dataset above, and equivalence
classes comprising \skad\ and \stew, and everybody else:

<<>>=
ec <-c(autimatic=1, RUSH=1, Skadoodle=2, Stewie2K=2, tarik=1)
@

@@

Find the MLE for $\lambda$:

@@

consider Table~\ref{kka}, originally considered
by~\citet{hankin2020}.  Considering only non-drawn games for the
moment, we wish to represent the first-mover advantage using a ``White
monster'', of Bradley-Terry strength \white

\begin{table}
\centering
\begin{tabular}{@{} *{4}{c} @{}}
\headercell{\small{Plays White,}\\ \small{wins}} & \multicolumn{3}{c@{}}{\small{Plays black, loses}}\\
\cmidrule(l){2-4}
& An &  Kr & Ks        \\
\midrule
  An & -  &  18 &   6  \\
  Kr &  7 &  -  &  18  \\
  Ks & 15 &  30 &  -   \\
\end{tabular}\quad
\begin{tabular}{@{} *{4}{c} @{}}
\headercell{\small{Plays W,}\\ \small{draws}} & \multicolumn{3}{c@{}}{\small{Plays black, draws}}\\
 \cmidrule(l){2-4}
& An &  Kr & Ks        \\
\midrule
  An & -  &  20 &  20  \\
  Kr & 29 &  -  &  72  \\
  Ks & 26 &  57 &  -   \\
\end{tabular}\quad
\begin{tabular}{@{} *{4}{c} @{}}
\headercell{\small{Plays W,}\\ \small{loses}} & \multicolumn{3}{c@{}}{\small{Plays black, wins}}\\
\cmidrule(l){2-4}
& An &  Kr & Ks        \\
\midrule
  An & -  &   5 &   11  \\
  Kr & 13 &  -  &    9  \\
  Ks &  2 &   7 &  -   \\
\end{tabular}
\caption{Results of 365 chess games \label{kka} between Grandmaster
  Anand, Grandmaster Karpov, and Grandmaster Kasparov.  For example,
  Anand played white against Karpov a total of $18+20+13=51$ times and
  won 18, drew 20, lost 13}
\end{table}

\begin{table}
\centering
\begin{tabular}{l|cc|cc}\\&
  \multicolumn{2}{c}{$i$ plays at home} &
  \multicolumn{2}{c}{$i$ plays away} \\
 & Reified & $\lambda$ &
   Reified & $\lambda$ \\
  \hline
  $\Prob{(\mbox{$i$ wins})}$  &
  $\frac{p_i+\home\oo}{p_i+p_j+\home}$ &
  $\frac{\lambda p_i}{\lambda p_i+p_j}$ &
  $\frac{p_i}{p_i+p_j+\home}$ &
  $\frac{p_i}{\lambda p_i+p_j}$ \\
  $\Prob{(\mbox{$i$ loses})}$ &
  $\frac{p_j\oo}{p_i+p_j+\home}$   &
  $\frac{p_j}{p_i+\lambda p_j}$  &
  $\frac{p_j+\home}{p_i+p_j+\home}$ &
  $\frac{\lambda p_j}{p_i + \lambda p_j}$
  \\
\end{tabular}
\caption{Probabilities of a win, draw, loss for the two teams under a reified
  Bradley-Terry model\label{probabilities} and a lambda thing}
\end{table}

>>>>

One generalization of the Bradley-Terry model is due
to~\citet{luce1959}, in which the probability of competitor~$i$
winning in a field of~$j=1,\ldots, n$ is $\frac{p_i}{p_1+\cdots
  +p_n}$.  It is then natural to quantify some ephemeral enhancement
to competitor~$i$ using a multiplicative factor $\lambda$.  Then
$\Prob(\mbox{$i$ wins})=\frac{\lambda p_i}{p_1+\cdots+\lambda
  p_i+\cdots+p_n}$.  Examples of such enhancements include the
home-ground advantage in football and rugby, or playing white in
chess~\citep{hankin2020}.  Note that $\lambda < 1$ would correspond to
a decrease in competitive ability.

Further, one might follow~\cite{hankin2017_rnw} and consider teams of
players.  In doubles tennis, for example, $\left\lbrace
p_1,p_2\right\rbrace$ might play $\left\lbrace p_3,p_4\right\rbrace$
and win with probability
$\frac{p_1+p_2}{p_1+p_2+p_3+p_4}$~\citep{hankin2010}.  It is natural to
combine this idea with that of a multiplicative factor.  For example,
suppose $p_1$ has a special move against $p_3$, and $p_4$ has just
been given a performance-enhancing drug, we might have a likelihood
functions proportional to $\frac{\lambda p_1+ p_2}{\lambda
  p_1+p_2+p_3+\mu p_4}$.  Then the effectiveness of $p_3$'s special
move and the enhancement afforded by the drug are measured by
$\lambda$ and $\mu$ respectively with values $>1$ corresponding to a
positive effect.

@@

In {\tt hyper3} idiom, a term such as $\lambda p_R + p_C$ is
represented as a named numeric vector.  For example, if the home
strength were $\lambda = 1.8$ then the first bracket would be {\tt
  c(Rangers=1.8, Celtic=1)} with an associated power of $-81$.  The
computational methodology is a generalization of that used
in~\cite{hankin2024_hyper3} in which the multiplicative factors are
known non-negative integers.

@@

Further, we may calculate a support function over the
$(\lambda,D)$-plane, shown in figure~\ref{contourlambdaD}.  For
example, we may test the hypothesis that $\lambda=1$ [corresponding to
  the right hand side of figure~\ref{contourlambdaD}] for which the
support is about \Sexpr{round(f(maxv) - jj$objective,2)}, far
exceeding the two units of support per degree of freedom
criteron~\citep{edwards1992}.  Alternatively, we can observe that this
lies in the tail region of its asymptotic null $\chi^2_4$ distribution
with a $p$-value of about \Sexpr{signif(pchisq(f(maxv) -
  jj$objective,lower.tail=FALSE,df=2),digits=2)}.

@@

The probability model of Figure~\ref{tikzabcde} is applied, in a
simple one-parameter form, to two datasets: one a set of preferences
of different types of sushi, and one in the field of e-sports.

@@

The red bus-blue bus problem was originally stated by
\cite{chipman1960} and discussed by \cite{davidson1970} and
\cite[p153]{alvo2014}.  
by \cite{tverskey1972}.
The canonical problem is as follows. A traveller has a choice of going
to work by different transport methods.  Here, we can assume that the
traveller has strong but unknown views on different forms of transport
which might incorporate a random component.  If considering whether to
use a car or a bus, the colour of the bus is of no importance to him.
Such preferences are problematic for Bradley-Terry: If our subject is
indifferent between cars and buses, say, $p_{C} = p_{RB}$ and also
indifferent to colour of bus, so $p_{RB}=p_{BB}$, the following
paradox occurs.  Bradley-Terry strengths must be equal so $p_{C} =
p_{RB} = p_{BB}=\frac{1}{3}$.  Thus $\Prob(C| \left\lbrace
C,RB\right\rbrace)=\frac{1}{2}$.  Now we introduce a blue bus as an
option.  Any realistic choice model will have $\Prob(C | \left\lbrace
C,RB,BB\right\rbrace)=\frac{1}{2}$ but Bradley-Terry gives us
$\Prob(C| \left\lbrace C,RB\right\rbrace)=\frac{1}{3}$.  Various
solutions to this problem have been proposed.  \cite{tverskey1972},
for example, presents an approach involving overlapping aspects;

If we ask him to
rank his options, it is highly probable that he will put RB and BB
consecutively (because they are essentially indistinguishable).  Can
we quantify the strength of this effect?  To do this, we define a
bespoke function \code{RB\_BB\_LF()} which returns a \code{hyper3}
log-likelihood function corresponding to repeated observations of our
commuter's reported ranks for the five options:

@@

Easily satisfying \citeauthor{edwards1992}'s two-units-of-support
criterion; alternatively, we could observe that this observation lies
in the tail region of its asymptotic $\chi^2_1$ distribution (Wilks's
theorem), with a $p$-value of \Sexpr{signif(pchisq(suppdiff*2, df=1,
  lower.tail=FALSE),3)}.

@@

@@

@@

To assess this, we follow exactly the same protocol as for the sushi
preference dataset but using the deathorder statistic from the dataset
above.  We specify two equivalence classes: one comprising \skad\ and
\stew, and one everyone else.  Figure~\ref{profsupcounterstrike} gives
a support function for $\lambda$ showing, among other things, that we
may reject the hypothesis that $\lambda=1$ in favour of $\lambda<1$ by
a likelihood ratio of about $\Sexpr{round(support,2)}$.  We observe
that this exceeds the two units of support criterion of
\cite{edwards1992} [asymptotic $p$-value
  $\simeq\Sexpr{round(pvalue,3)}$] so we may infer that there is
indeed some systematic mechanism---perhaps a positive feedback
loop---which tends to magnify differences in death statistic.
